A highly-available move operation for replicated trees

### 2020-03-03 paper/move-op.tex
Revert "Reformat move-op paper"
This reverts commit 6298d2abba36097ab8ba547704cfbe9cc04d5804.

\documentclass{vldb}

\usepackage[hidelinks]{hyperref}
\usepackage{textcomp} % \textmu
\usepackage{booktabs} % \toprule etc.
\usepackage{setspace} % \setstretch
\usepackage{balance}
\usepackage{tikz}
\usetikzlibrary{arrows.meta}

\urlstyle{rm}

@@

\vldbTitle{A highly-available move operation for replicated trees}
\vldbAuthors{Martin Kleppmann, Dominic P. Mulligan, Victor B. F. Gomes, and Alastair R. Beresford}
\vldbDOI{https://doi.org/10.14778/xxxxxxx.xxxxxxx}
\vldbVolume{13}
\vldbNumber{xxx}
\vldbYear{2020}

@@

\numberofauthors{4}
\author{
\alignauthor
    Martin Kleppmann\\
    \affaddr{University of Cambridge}\\
    \affaddr{Cambridge, UK}\\
    \email{mk428@cl.cam.ac.uk}
\alignauthor
    Dominic P.\ Mulligan\\
    \affaddr{Arm Research}\\
    \affaddr{Cambridge, UK}\\
    \email{Dominic.Mulligan@arm.com}
\alignauthor
    Victor B.\ F.\ Gomes\\
    \affaddr{University of Cambridge}\\
    \affaddr{Cambridge, UK}\\
    \email{vb358@cl.cam.ac.uk}
\and\alignauthor
    Alastair R.\ Beresford\\
    \affaddr{University of Cambridge}\\
    \affaddr{Cambridge, UK}\\
    \email{arb33@cl.cam.ac.uk}
}
\maketitle

@@

@@

  \node [rotate=90,blue,font=\scriptsize] at (7.5,1.5) {network communication};

@@

  \node [fill=red!15] at (9.63,4.10) {(\emph{a})};
  \node [fill=red!15] at (13.98,4.10) {(\emph{b})};
  \node [fill=red!15] at (9.64,1.10) {(\emph{c})};
  \node [fill=red!15] at (13.95,1.10) {(\emph{d})};
  \node [rectangle,draw] at (10.7,3) {

@@

  \node [rectangle,draw] at (13.3,3) {

@@

  \node [rectangle,draw] at (10.5,0) {

@@

  \node [rectangle,draw] at (13.1,0) {

@@

@@

\begin{figure*}
\centering
\begin{tikzpicture}
  \tikzstyle{time}=[thick,->,gray]
  \tikzstyle{network}=[thick,dashed,blue,-{Stealth[length=3mm]}]
  \node [anchor=east] at (-1.2,3) {Replica 1:};
  \node [anchor=east] at (-1.2,0) {Replica 2:};
  \node [rectangle,draw] (start1) at (0,3) {
      \begin{tikzpicture}[level distance=7mm]
      \node {$\mathsf{root}$} child {node {$A$} child {node {$C$}}} child {node {$B$}};
      \end{tikzpicture}
  };
  \node [rectangle,draw] (start2) at (0,0) {
      \begin{tikzpicture}[level distance=7mm]
      \node {$\mathsf{root}$} child {node {$A$} child {node {$C$}}} child {node {$B$}};
      \end{tikzpicture}
  };
  \node [rectangle,draw] (change1) at (4.5,3) {
      \begin{tikzpicture}[level distance=7mm]
      \node {$\mathsf{root}$} child {node {$A$} child {node {$B$}} child {node {$C$}}};
      \end{tikzpicture}
  };
  \node [rectangle,draw] (change2) at (4.5,0) {
      \begin{tikzpicture}[level distance=7mm]
      \node {$\mathsf{root}$} child {node {$B$} child {node {$A$} child {node {$C$}}}};
      \end{tikzpicture}
  };
  \node [rectangle,draw,inner sep=3mm] (merge1) at (8,3) {?};
  \node [rectangle,draw,inner sep=3mm] (merge2) at (8,0) {?};
  \draw [time] (start1.east) -- node [above,text width=2.5cm,text centered,inner ysep=5pt] {Move $B$ to be a child of $A$} (change1.west);
  \draw [time] (change1.east) -- (merge1.west);
  \draw [time] (start2.east) -- node [above,text width=2.5cm,text centered,inner ysep=5pt] {Move $A$ to be a child of $B$} (change2.west);
  \draw [time] (change2.east) -- (merge2.west);
  \draw [network] (6.0,0) to [out=90,in=270] (6.7,3);
  \draw [network] (6.0,3) to [out=270,in=90] (6.7,0);
  \node [rotate=90,blue,font=\scriptsize] at (7.0,1.5) {network communication};
  \path [draw,dotted] (-2.5,1.6) -- (8.5,1.6);
  %%%%%
  \node [rectangle,draw] at (10.75,3) {
      \begin{tikzpicture}[level distance=7mm]
      \useasboundingbox (-1.3,-1.3) rectangle (1,1.2);
      \node [fill=red!15] at (-1.07,1.02) {(\emph{a})};
      \node at (0,1) {$\mathsf{root}$};
      \node (a1) {$A$} child {node (b1) {$B$}} child {node {$C$}};
      \draw (b1.south west) .. controls (-2,-2) and (0,1.5) .. (a1.north);
      \end{tikzpicture}
  };
  \node [rectangle,draw] at (13.6,3) {
      \begin{tikzpicture}[level distance=7mm]
      \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
      \node [fill=red!15] at (0.93,0.02) {(\emph{b})};
      \tikzstyle{level 1}=[sibling distance=12mm]
      \tikzstyle{level 2}=[sibling distance=6mm]
      \node {$\mathsf{root}$}
          child {node {$A$} child {node {$B$}} child {node {$C$}}}
          child {node {$B'$} child {node {$A'$} child {node {$C'$}}}};
      \end{tikzpicture}
  };
  \node [rectangle,draw] (start) at (10.75,0) {
      \begin{tikzpicture}[level distance=7mm]
      \useasboundingbox (-1.17,-2.3) rectangle (1.15,0.2);
      \node [fill=red!15] at (-0.95,0.03) {(\emph{c})};
      \node {$\mathsf{root}$} child {node {$A$} child {node {$B$}} child {node {$C$}}};
      \end{tikzpicture}
  };
  \node [rectangle,draw] (right) at (13.6,0) {
      \begin{tikzpicture}[level distance=7mm]
      \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
      \node [fill=red!15] at (0.91,0.03) {(\emph{d})};
      \node {$\mathsf{root}$} child {node {$B$} child {node {$A$} child {node {$C$}}}};
      \end{tikzpicture}
  };
\end{tikzpicture}
\caption{Initially, nodes $A$ and $B$ are siblings. Replica 1 moves $B$ to be a child of $A$, while concurrently replica 2 moves $A$ to be a child of $B$. Boxes (\emph{a}) to (\emph{d}) show possible outcomes after the replicas have communicated and merged their states.}\label{fig:move-cycle}
\end{figure*}

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth,keepaspectratio=true]{gdrive-error.png}
  \caption{Error message produced by Google Drive Backup and Sync on Mac OS as a result of performing the operations shown in Figure~\ref{fig:move-cycle}, indicating a bug in the underlying replication algorithm.}
  \label{fig:gdrive-error}
\end{figure*}

@@

\internallinenumbers\setstretch{1.2}

@@

The \isa{apply\_op} function (lines 43--51) takes two arguments: a \isa{Move} operation to apply and the current replica state; and it returns the new replica state.
The \isa{'t::\{linorder\}} constraint in the type signature indicates that timestamps \isa{'t} are instances of \isa{linorder} type class, and they can therefore be compared with the \isa{<} operator defining a linear (or total) order.
This comparison occurs on line 49.

@@

We now discuss the correctness properties of the algorithm from \S\ref{sec:algorithm}.
All theorems stated here have been formally proved and mechanically checked using Isabelle.
For space reasons we give only a brief sketch of the proofs in this paper; the Isabelle files containing the full details are open source~\cite{SourceFiles}.

@@

\noindent The \isa{\isacharat} operator concatenates two lists.
Let \isa{t1 = move\_time oper1} and \isa{t2 = move\_time oper2} be the timestamps of the two operations \isa{oper1} and \isa{oper2}.
Assume \isa{t1 < t2} without loss of generality.
We can now prove lemma \isa{apply\_op\_commute2} by induction over \isa{log}.

@@

For the inductive step, let the log be \isa{logop {\isacharhash} ops} (where \isa{\isacharhash} is the \emph{list cons} operator).
We now need to show that:
\begin{isabelle}
apply\_op oper2 (apply\_op oper1 (logop {\isacharhash} ops, tree)) =\\
apply\_op oper1 (apply\_op oper2 (logop {\isacharhash} ops, tree)).
\end{isabelle}
Let \isa{t3 = log\_time logop} be the timestamp of the operation at the head of the log.
We now consider three cases:
\begin{enumerate}
    \item \isa{t3 < t1 < t2}.\\
        Then neither applying \isa{oper1} nor applying \isa{oper2} will undo \isa{logop}.
        Thus, the two operations are applied in the same way as in the base case.
    \item \isa{t1 < t3 < t2}. Then\\
        \isa{apply\_op oper2 (apply\_op oper1 (logop {\isacharhash} ops, tree))}\\first undoes \isa{logop}, then applies \isa{oper1} using \isa{do\_op}, then reapplies \isa{logop} using \isa{redo\_op}, and finally applies \isa{oper2} using \isa{do\_op}.
        On the other hand,\\\isa{apply\_op oper1 (apply\_op oper2 (logop {\isacharhash} ops, tree))}\\first applies \isa{oper2} using \isa{do\_op}, then undoes it again, then undoes \isa{logop}, applies \isa{oper1} and reapplies \isa{logop} as before, and finally reapplies \isa{oper2} using \isa{redo\_op}.
        The end result is the same.
    \item \isa{t1 < t2 < t3}.\\
        Then both applying \isa{oper1} and applying \isa{oper2} first undoes \isa{logop}, yielding \isa{tree' = undo\_op (logop, tree)}.
        From the inductive hypothesis we have:
        \begin{isabelle}
        apply\_op oper2 (apply\_op oper1 (ops, tree')) =\\
        apply\_op oper1 (apply\_op oper2 (ops, tree'))
        \end{isabelle}
        We now apply \isa{oper1} and \isa{oper2} as in the base case, and finally reapply \isa{logop} to both sides using \isa{redo\_op}, resulting in the same state on both sides.
\end{enumerate}

@@

Gomes et al.~\cite{Gomes:2017gy} define a framework in Isabelle/HOL for proving the strong eventual consistency properties of CRDTs above an axiomatic network model.
Using our commutativity proof above we integrate our tree datatype with this framework, and thus prove that our move operation on trees does indeed guarantee strong eventual consistency.
The details appear in our Isabelle theory files~\cite{SourceFiles}.

@@

That is, if \isa{executable\_apply\_ops} and \isa{apply\_ops} are applied to the same list of operations, they produce identical logs, and also produce trees that contain the same set of key-value bindings---i.e. the trees are extensionally equivalent, despite having very different in-memory representations.
We can also prove corollaries of the theorems \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic} for the hash-map-based implementation.
%The analogue of \isa{apply\_ops\_unique\_parent} holds trivially since by construction, the hash-map contains at most one (metadata, parent) pair for a given child.

@@

\begin{table}
  \caption{Median response times of an RPC request between replicas in different regions.}
  \label{tab:rpc-times}
  \vspace{4pt}
  \begin{tabular}{r|rrr}
    \toprule
                    & to California & to Ireland & to Singapore \\
    \midrule
    from California & ---           & 139 ms     & 175 ms       \\
    from Ireland    & 145 ms        & ---        & 182 ms       \\
    from Singapore  & 176 ms        & 178 ms     & ---          \\
    \bottomrule
\end{tabular}
\end{table}

\begin{figure}
  \includegraphics[width=\columnwidth,keepaspectratio=true]{plot.pdf}
  \caption{Median execution time for applying an operation to the replica state, on an Amazon EC2 c5.large instance, at varying operation throughput rates. Error bars indicate the minimum and 95th percentile execution times.}
  \label{fig:plots}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth,keepaspectratio=true]{leader-plot.pdf}
  \caption{Median time to apply a local CRDT operation compared to the median time to perform a move operation using state machine replication, with a leader located in another region. Note the log scale on both axes.}
  \label{fig:leader-plot}
\end{figure}

@@

Many replicated data systems use optimistic replication \cite{Saito:2005jw}, which allows the state of replicas to temporarily diverge, in order to achieve better performance and availability in the presence of faults than strongly consistent systems~\cite{Bailis:2014th,Gilbert:2002il}.
As a consequence, these systems require a mechanism for merging or reconciling conflicting updates that were made concurrently on different replicas.
For example, version control systems such as Git~\cite{Chacon:2014kr} leave conflicts for the user to resolve manually.
Databases such as Dynamo~\cite{DeCandia:2007ui} and Bayou~\cite{Terry:1995dn} rely on the application programmer to provide explicit conflict resolution logic; however, such logic is difficult to get right \cite{Bailis:2013jc,Burckhardt:2014hy,Gomes:2017gy}.
Hence, we focus on systems that automatically ensure that all replicas converge towards a consistent state, without requiring custom application logic---a consistency model known as \emph{strong eventual consistency}~\cite{Shapiro:2011un,Gomes:2017gy}.

@@

Our algorithm is an operation-based Conflict-free Replicated Data Type or CRDT~\cite{Shapiro:2011wy,Shapiro:2011un,Burckhardt:2014ft}.
All CRDTs share the property that concurrent changes on different replicas can be merged in any order; any two replicas that have seen the same set of updates are guaranteed to be in the same state, regardless of the order in which they processed these updates.
Several CRDTs for trees have been proposed:
\begin{itemize}
    \item Martin et al.~\cite{Martin:2010ih} define a CRDT for XML data, and Kleppmann and Beresford~\cite{Kleppmann:2016ve} define a CRDT with a JSON data model.
        However, these algorithms only deal with insertion and deletion of tree nodes, and do not support moves.
        A move operation can be emulated by deleting and re-inserting the moved node, but this approach suffers from the duplication problem demonstrated in \S\ref{sec:move-same-item}.
    \item As discussed in \S\ref{sec:impossibility}, Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} propose two implementations for a replicated filesystem: a CRDT in which conflicting moves are handled by duplicating tree nodes (as in Figures~\ref{fig:move-same-item}b and~\ref{fig:move-cycle}b), and a centralised implementation in which move operations must obtain a lock before executing (not a CRDT since it relies on synchronous coordination).
    \item Ahmed-Nacer et al.~\cite{AhmedNacer:2012us} outline approaches to handling conflicts on trees, but do not provide algorithms.
    \item Tao et al.~\cite{Tao:2015gd} propose handling conflicting move operations by allowing the same object to appear in more than one location; thus, their datatype is strictly a DAG, not a tree.
        Some conflicts are handled by duplicating tree nodes.
        Tao et al.\ also perform experiments with Dropbox, Google Drive, and OneDrive, similar to our experiments discussed in \S\ref{sec:move-is-hard}.
\end{itemize}

@@

\section*{Acknowledgements}

The authors wish to acknowledge the support of The Boeing Company and the EPSRC ``REMS: Rigorous Engineering for Mainstream Systems'' programme grant (EP/K008528).
Martin Kleppmann is supported by a Leverhulme Early Career Fellowship and by the Isaac Newton Trust.
Our evaluation was conducted using AWS credits from the AWS Educate program.
Thank you to Marc Shapiro for feedback on a draft of this paper.

%\balance
\bibliographystyle{abbrv}
\bibliography{references}{}
\end{document}

>>>>

\documentclass[sigplan,anonymous]{acmart}

% EuroSys recommended formatting
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

\usepackage{tikz}
\usetikzlibrary{arrows.meta}

@@

\setcopyright{rightsretained}
\copyrightyear{2019}
\acmYear{2019}
\acmDOI{}
\acmConference[]{}{}{}
\acmBooktitle{}
\acmPrice{}
\acmISBN{}

@@

\author{Martin Kleppmann}
\email{mk428@cl.cam.ac.uk}
\orcid{0000-0001-7252-6958}
\affiliation{%
  \institution{Department of Computer Science and Technology}
  \streetaddress{15 JJ Thomson Avenue}
  \city{Cambridge}
  \state{}
  \postcode{CB3 0FD}
  \country{United Kingdom}
}

\author{Dominic P.\ Mulligan}
\email{Dominic.Mulligan@arm.com}
\orcid{0000-0003-4643-3541}
\affiliation{%
  \institution{Arm Research}
  \streetaddress{110 Fulbourn Road}
  \city{Cambridge}
  \state{}
  \postcode{CB1 9NJ}
  \country{United Kingdom}
}

\author{Victor B.\ F.\ Gomes}
\email{vb358@cl.cam.ac.uk}
\orcid{0000-0002-2954-4648}
\affiliation{%
  \institution{Department of Computer Science and Technology}
  \streetaddress{15 JJ Thomson Avenue}
  \city{Cambridge}
  \state{}
  \postcode{CB3 0FD}
  \country{United Kingdom}
}

\author{Alastair R.\ Beresford}
\email{arb33@cl.cam.ac.uk}
\orcid{0000-0003-0818-6535}
\affiliation{%
  \institution{Department of Computer Science and Technology}
  \streetaddress{15 JJ Thomson Avenue}
  \city{Cambridge}
  \state{}
  \postcode{CB3 0FD}
  \country{United Kingdom}
}

%\renewcommand{\shortauthors}{Kleppmann, Mulligan, Gomes, and Beresford}

@@

\settopmatter{printfolios=true} % Recommended EuroSys formatting
\maketitle

@@

  \node [rotate=90,blue,font=\footnotesize] at (7.5,1.5) {network communication};

@@

  \node [fill=red!15] at (9.83,4.10) {(\emph{a})};
  \node [fill=red!15] at (14.18,4.10) {(\emph{b})};
  \node [fill=red!15] at (9.84,1.10) {(\emph{c})};
  \node [fill=red!15] at (14.15,1.10) {(\emph{d})};
  \node [rectangle,draw] at (10.9,3) {

@@

  \node [rectangle,draw] at (13.5,3) {

@@

  \node [rectangle,draw] at (10.7,0) {

@@

  \node [rectangle,draw] at (13.3,0) {

@@

\Description{}

@@

\begin{figure*}
\centering
\begin{tikzpicture}
  \tikzstyle{time}=[thick,->,gray]
  \tikzstyle{network}=[thick,dashed,blue,-{Stealth[length=3mm]}]
  \node [anchor=east] at (-1.2,3) {Replica 1:};
  \node [anchor=east] at (-1.2,0) {Replica 2:};
  \node [rectangle,draw] (start1) at (0,3) {
      \begin{tikzpicture}[level distance=7mm]
      \node {$\mathsf{root}$} child {node {$A$} child {node {$C$}}} child {node {$B$}};
      \end{tikzpicture}
  };
  \node [rectangle,draw] (start2) at (0,0) {
      \begin{tikzpicture}[level distance=7mm]
      \node {$\mathsf{root}$} child {node {$A$} child {node {$C$}}} child {node {$B$}};
      \end{tikzpicture}
  };
  \node [rectangle,draw] (change1) at (4.5,3) {
      \begin{tikzpicture}[level distance=7mm]
      \node {$\mathsf{root}$} child {node {$A$} child {node {$B$}} child {node {$C$}}};
      \end{tikzpicture}
  };
  \node [rectangle,draw] (change2) at (4.5,0) {
      \begin{tikzpicture}[level distance=7mm]
      \node {$\mathsf{root}$} child {node {$B$} child {node {$A$} child {node {$C$}}}};
      \end{tikzpicture}
  };
  \node [rectangle,draw,inner sep=3mm] (merge1) at (8,3) {?};
  \node [rectangle,draw,inner sep=3mm] (merge2) at (8,0) {?};
  \draw [time] (start1.east) -- node [above,text width=2.5cm,text centered,inner ysep=5pt] {Move $B$ to be a child of $A$} (change1.west);
  \draw [time] (change1.east) -- (merge1.west);
  \draw [time] (start2.east) -- node [above,text width=2.5cm,text centered,inner ysep=5pt] {Move $A$ to be a child of $B$} (change2.west);
  \draw [time] (change2.east) -- (merge2.west);
  \draw [network] (6.0,0) to [out=90,in=270] (6.7,3);
  \draw [network] (6.0,3) to [out=270,in=90] (6.7,0);
  \node [rotate=90,blue,font=\footnotesize] at (7.0,1.5) {network communication};
  \path [draw,dotted] (-2.5,1.6) -- (8.5,1.6);
  %%%%%
  \node [rectangle,draw] at (10.95,3) {
      \begin{tikzpicture}[level distance=7mm]
      \useasboundingbox (-1.3,-1.3) rectangle (1,1.2);
      \node [fill=red!15] at (-1.10,1.03) {(\emph{a})};
      \node at (0,1) {$\mathsf{root}$};
      \node (a1) {$A$} child {node (b1) {$B$}} child {node {$C$}};
      \draw (b1.south west) .. controls (-2,-2) and (0,1.5) .. (a1.north);
      \end{tikzpicture}
  };
  \node [rectangle,draw] at (13.8,3) {
      \begin{tikzpicture}[level distance=7mm]
      \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
      \node [fill=red!15] at (0.94,0.03) {(\emph{b})};
      \tikzstyle{level 1}=[sibling distance=12mm]
      \tikzstyle{level 2}=[sibling distance=6mm]
      \node {$\mathsf{root}$}
          child {node {$A$} child {node {$B$}} child {node {$C$}}}
          child {node {$B'$} child {node {$A'$} child {node {$C'$}}}};
      \end{tikzpicture}
  };
  \node [rectangle,draw] (start) at (10.95,0) {
      \begin{tikzpicture}[level distance=7mm]
      \useasboundingbox (-1.17,-2.3) rectangle (1.15,0.2);
      \node [fill=red!15] at (-0.98,0.03) {(\emph{c})};
      \node {$\mathsf{root}$} child {node {$A$} child {node {$B$}} child {node {$C$}}};
      \end{tikzpicture}
  };
  \node [rectangle,draw] (right) at (13.8,0) {
      \begin{tikzpicture}[level distance=7mm]
      \useasboundingbox (-1.15,-2.3) rectangle (1.15,0.2);
      \node [fill=red!15] at (0.93,0.03) {(\emph{d})};
      \node {$\mathsf{root}$} child {node {$B$} child {node {$A$} child {node {$C$}}}};
      \end{tikzpicture}
  };
\end{tikzpicture}
\caption{Initially, nodes $A$ and $B$ are siblings. Replica 1 moves $B$ to be a child of $A$, while concurrently replica 2 moves $A$ to be a child of $B$. Boxes (\emph{a}) to (\emph{d}) show possible outcomes after the replicas have communicated and merged their states.}\label{fig:move-cycle}
\Description{Visualisation of trees with concurrent move operations}
\end{figure*}

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth,keepaspectratio=true]{gdrive-error.png}
  \caption{Error message produced by Google Drive Backup and Sync on Mac OS as a result of performing the operations shown in Figure~\ref{fig:move-cycle}, indicating a bug in the underlying replication algorithm.}
  \label{fig:gdrive-error}
  \Description{Screenshot showing the errors ``Can't remove files with changes pending. Try again after sync is complete.'' and ``An unknown error occurred.''}
\end{figure*}

@@

\internallinenumbers

@@

The \isa{apply\_op} function (lines 43--51) takes two arguments: a \isa{Move} operation to apply and the current replica state; and it returns the new replica state.
The constraint \isa{'t::\{linorder\}} in the type signature indicates that timestamps \isa{'t} are instances of \isa{linorder} type class, and they can therefore be compared with the \isa{<} operator defining a linear (or total) order.
This comparison occurs on line 49.

@@

We now discuss the correctness properties of the algorithm from \S\ref{sec:algorithm}.
All theorems stated here have been formally proved and mechanically checked using Isabelle.
For space reasons we give only a brief sketch of the proofs in this paper; the Isabelle files containing the full details are open source [URL omitted for blind review].

% TODO \cite{SourceFiles}

@@

\noindent Here the \isa{\isacharat} operator concatenates two lists.
Let \isa{t1 = move\_time oper1} and \isa{t2 = move\_time oper2} be the timestamps of the two operations \isa{oper1} and \isa{oper2}.
Assume \isa{t1 < t2} without loss of generality.
We can now prove lemma \isa{apply\_op\_commute2} by induction over \isa{log}.

@@

For the inductive step, let the log be \isa{logop {\isacharhash} ops} (where \isa{\isacharhash} is the \emph{list cons} operator).
We now need to show that:
\begin{isabelle}
apply\_op oper2 (apply\_op oper1 (logop {\isacharhash} ops, tree)) =\\
apply\_op oper1 (apply\_op oper2 (logop {\isacharhash} ops, tree)).
\end{isabelle}
Let \isa{t3 = log\_time logop} be the timestamp of the operation at the head of the log.
We now consider three cases:
\begin{enumerate}
    \item \isa{t3 < t1 < t2}.
        Then neither applying \isa{oper1} nor applying \isa{oper2} will undo \isa{logop}.
        Thus, the two operations are applied in the same way as in the base case.
    \item \isa{t1 < t3 < t2}.
        Then \isa{apply\_op oper2 (apply\_op oper1 (logop {\isacharhash} ops, tree))} first undoes \isa{logop}, then applies \isa{oper1} using \isa{do\_op}, then reapplies \isa{logop} using \isa{redo\_op}, and finally applies \isa{oper2} using \isa{do\_op}.
        On the other hand, \isa{apply\_op oper1 (apply\_op oper2 (logop {\isacharhash} ops, tree))} first applies \isa{oper2} using \isa{do\_op}, then undoes it again, then undoes \isa{logop}, applies \isa{oper1} and reapplies \isa{logop} as before, and finally reapplies \isa{oper2} using \isa{redo\_op}.
        The end result is the same.
    \item \isa{t1 < t2 < t3}.
        Then both applying \isa{oper1} and applying \isa{oper2} first undoes \isa{logop}, yielding \isa{tree' = undo\_op (logop, tree)}.
        From the inductive hypothesis we have:
        \begin{isabelle}
        apply\_op oper2 (apply\_op oper1 (ops, tree')) =\\
        apply\_op oper1 (apply\_op oper2 (ops, tree'))
        \end{isabelle}
        We now apply \isa{oper1} and \isa{oper2} as in the base case, and finally reapply \isa{logop} to both sides using \isa{redo\_op}, resulting in the same state on both sides.
\end{enumerate}

@@

Gomes et al.~\cite{Gomes:2017gy} define a framework in Isabelle/HOL for proving the strong eventual consistency properties of CRDTs above an axiomatic network model.
Using our commutativity proof above we integrate our tree datatype with this framework, and thus prove that our move operation on trees does indeed guarantee strong eventual consistency.
The details appear in our Isabelle theory files [URL omitted for blind review]. % \cite{SourceFiles}.

@@

That is, if \isa{executable\_apply\_ops} and \isa{apply\_ops} are applied to the same list of operations, they produce identical logs, and also produce trees that contain the same set of key-value bindings---i.e. the trees are extensionally equivalent, despite having very different in-memory representations.
We can also prove corollaries of \isa{apply\_ops\_commutes} and \isa{apply\_ops\_acyclic} for the hash-map-based implementation.
%The analogue of \isa{apply\_ops\_unique\_parent} holds trivially since by construction, the hash-map contains at most one (metadata, parent) pair for a given child.

@@

\begin{table}
  \caption{Median response times of an RPC request between replicas in different regions.}
  \label{tab:rpc-times}
  \begin{tabular}{r|rrr}
    \toprule
                    & to California & to Ireland & to Singapore \\
    \midrule
    from California & ---           & 139 ms     & 175 ms       \\
    from Ireland    & 145 ms        & ---        & 182 ms       \\
    from Singapore  & 176 ms        & 178 ms     & ---          \\
    \bottomrule
\end{tabular}
\end{table}

\begin{figure}
  \includegraphics[width=\columnwidth,keepaspectratio=true]{plot.pdf}
  \caption{Median execution time for applying an operation to the replica state, on an Amazon EC2 c5.large instance, at varying operation throughput rates. Error bars indicate the minimum and 95th percentile execution times.}
  \label{fig:plots}
  \Description{Plot in which the x axis goes from 0 to 600 move operations per second, and the y axis goes from 0 to 2 milliseconds. Local operations take a near-constant time around 50 microseconds; the time for remote operations increases linearly from 0.2ms to 1.4ms.}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth,keepaspectratio=true]{leader-plot.pdf}
  \caption{Median time to apply a local CRDT operation compared to the median time to perform a move operation using state machine replication, with a leader located in another region. Note the log scale on both axes.}
  \label{fig:leader-plot}
  \Description{Local CRDT operations take about 50 microseconds, and reach saturation at 600 ops/sec. Leader-based operations take about 150--180ms, and reach saturation at 14,000 ops/sec.}
\end{figure}

@@

Many replicated data systems use optimistic replication~\cite{Saito:2005jw}, which allows the state of replicas to temporarily diverge, in order to achieve better performance and availability in the presence of faults than strongly consistent systems~\cite{Bailis:2014th,Gilbert:2002il}.
As a consequence, these systems require a mechanism for merging or reconciling conflicting updates that were made concurrently on different replicas.
For example, version control systems such as Git~\cite{Chacon:2014kr} leave conflicts for the user to resolve manually.
Databases such as Dynamo~\cite{DeCandia:2007ui} and Bayou~\cite{Terry:1995dn} rely on the application programmer to provide explicit conflict resolution logic; however, such logic is difficult to get right \cite{Bailis:2013jc,Burckhardt:2014hy,Gomes:2017gy}.
Hence, we focus on systems that automatically ensure that all replicas converge towards a consistent state, without requiring custom application logic---a consistency model known as \emph{strong eventual consistency}~\cite{Shapiro:2011un,Gomes:2017gy}.

@@

Our algorithm is an example of an operation-based Conflict-free Replicated Data Type or CRDT~\cite{Shapiro:2011wy,Shapiro:2011un,Burckhardt:2014ft}.
All CRDTs share the property that concurrent changes on different replicas can be merged in any order; any two replicas that have seen the same set of updates are guaranteed to be in the same state, regardless of the order in which they processed these updates.
Several CRDTs for trees have been proposed:
\begin{itemize}
    \item Martin et al.~\cite{Martin:2010ih} define a CRDT for XML data, and Kleppmann and Beresford~\cite{Kleppmann:2016ve} define a CRDT for JSON.
        However, these algorithms only deal with insertion and deletion of tree nodes, and do not support moves.
        A move operation can be emulated by deleting and re-inserting the moved node, but this approach suffers from the duplication problem demonstrated in \S\ref{sec:move-same-item}.
    \item As discussed in \S\ref{sec:impossibility}, Najafzadeh et al.~\cite{Najafzadeh:2017vk,Najafzadeh:2018bw} propose two implementations for a replicated filesystem: a CRDT in which conflicting moves are handled by duplicating tree nodes (as in Figures~\ref{fig:move-same-item}b and~\ref{fig:move-cycle}b), and a centralised implementation in which move operations must obtain a lock before executing (not a CRDT since it relies on synchronous coordination).
    \item Ahmed-Nacer et al.~\cite{AhmedNacer:2012us} outline approaches to handling conflicts on trees, but do not provide algorithms.
    \item Tao et al.~\cite{Tao:2015gd} propose handling conflicting move operations by allowing the same object to appear in more than one location; thus, their datatype is strictly a DAG, not a tree.
        Some conflicts are handled by duplicating tree nodes.
        Tao et al.\ also perform experiments with Dropbox, Google Drive, and OneDrive, similar to our experiments discussed in \S\ref{sec:move-is-hard}.
\end{itemize}

@@

\begin{acks}
The authors wish to acknowledge the support of The Boeing Company and the EPSRC ``REMS: Rigorous Engineering for Mainstream Systems'' programme grant (EP/K008528).
Martin Kleppmann is supported by a Leverhulme Trust Early Career Fellowship and by the Isaac Newton Trust.
Our evaluation was conducted using AWS credits from the AWS Educate program.
Thank you to Marc Shapiro for feedback on a draft of this paper.
\end{acks}

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}{}
\end{document}
